---
title: "Gene-level and Global Assessment of Unwanted Variation in the Recount3 Database"
output: html_document
date: "2025-12-12"
editor_options: 
  chunk_output_type: console
---

# Required Libraries and Packages

The following R packages are required to reproduce this analysis.

```{r}
library(SummarizedExperiment)
library(matrixTests)
library(matrixStats)
library(pbmcapply)
library(recount3)
library(parallel)
library(ggplot2)
library(mclust)
library(purrr)
library(edgeR)
library(Rfast)
library(irlba)
library(qs)
```

## Downloading the Recount3 Database

All available human RNA-seq projects are retrieved from the Recount3 database
using `available_projects()`. Projects with fewer than 20 samples are excluded
to ensure sufficient statistical power for downstream assessments. The
remaining studies are downloaded and saved as compressed `.qs` files for
efficient storage and retrieval. A fault-tolerant wrapper (`safely()` from
`purrr`) is used to handle individual download failures without interrupting
the full pipeline.

```{r}
human.projects <- available_projects()
# 2025-11-21 11:45:36.024052 caching file sra.recount_project.MD.gz.
# 2025-11-21 11:45:36.629394 caching file gtex.recount_project.MD.gz.
# 2025-11-21 11:45:37.174104 caching file tcga.recount_project.MD.gz.

dim(human.projects) # 8742  6
length(unique(human.projects$project)) # 8742
human.projects <- human.projects[human.projects$n_samples >= 20, ] # 2555    6
human.projects <- human.projects[order(human.projects$n_samples, decreasing = TRUE) , ]
sum(human.projects$n_samples) # 297113
lapply(
  1:100,
  function(x) {
    tryCatch({
      data.id <- gsub("_", " ", human.projects$project[x])
      file.name <- paste0("Recount3_", data.id, ".qs")
      message("Processing: ", file.name)
      if (!file.name %in% list.files()) {
        se.obj <- create_rse(project.info[x, ])
        qs::qsave(se.obj, file.name)
      }
      return(TRUE)
    }, error = function(e) {
      message("Error in index ", x, ": ", conditionMessage(e))
      return(NULL) # keep going
    })
  }
)

safe.fun <- safely(function(x) {
  data.id <- gsub("_", " ", human.projects$project[x])
  file.name <- paste0("Recount3_", data.id, ".qs")
  message("Processing: ", file.name)
  if (!file.name %in% list.files()) {
    se.obj <- create_rse(human.projects[x, ])
    qs::qsave(se.obj, file.name)
  }
  TRUE
})
results <- map(1:nrow(human.projects), safe.fun)
```

## Removing Studies with Fewer than 20 Samples

After downloading, studies are further filtered to retain only those with
exactly 63,856 genes (the expected number for the Recount3 human reference)
and more than 20 samples. This step ensures consistency in gene space across
all studies and removes studies too small for reliable statistical assessment.
A total of 2,284 studies comprising 221,333 samples are retained for
downstream analysis.

```{r}
setwd("/Users/molania.r/Documents/Projects/Project_DFCI_2_RUVprpsApplication/Application_RUVprps/Recount3/Data/")
all.studies <- list.files(pattern = '.qs') 
length(all.studies) == 2448 # RNA-seq studies

samples.genes <- mclapply(
    all.studies,
    function(x){
        temp.data <- qs::qread(x)
        data.frame(genes = nrow(temp.data), samples = ncol(temp.data))
    }, mc.cores = 14
    )
samples.genes <- do.call(rbind, samples.genes)
samples.genes$studies <- all.studies
samples.genes <- samples.genes[samples.genes$genes == '63856',  ]
samples.genes <- samples.genes[samples.genes$samples > 20,  ]
sum(samples.genes$samples) # 221,333 samples
all.studies <- samples.genes$studies # 2284 studies
```

### Library Size Effects on Gene Expression and PCA

To quantify the extent to which library size (total read count per sample)
constitutes a source of unwanted variation, two complementary metrics are
computed for each study. First, the proportion of genes whose expression is
correlated with library size ($|\rho| > 0.3$, Spearman) is calculated at the
gene level. Second, the variance in library size explained by the first two
principal components ($R^2$) is assessed at the global level. Samples
identified as outliers based on the interquartile range (IQR) of log~2~
library size are removed prior to analysis. Gene filtering is performed using
`edgeR::filterByExpr()` to retain only adequately expressed protein-coding
genes.

```{r}
genes.pca.ls <- function(file) {
  withCallingHandlers(
    expr = tryCatch({
      temp.data <- qs::qread(file)
      gene.info <- rowData(temp.data)
      raw.data <- assay(
          temp.data[gene.info$gene_type == 'protein_coding', ], 
          "raw_counts"
          )
      keep.genes <- edgeR::filterByExpr(y = raw.data)
      raw.data <- raw.data[keep.genes, , drop = FALSE]
      ls <- log2(colSums(raw.data))
      q1 <- quantile(ls, 0.25)
      q3 <- quantile(ls, 0.75)
      iqr <- q3 - q1
      lower <- q1 - 1.5 * iqr
      upper <- q3 + 1.5 * iqr
      outliers <- !ls < lower | ls > upper
      raw.data <- raw.data[ ,outliers]
      ls <- ls[outliers]
      cpm.norm <- edgeR::cpm(raw.data, log = TRUE)
      pca <- irlba::prcomp_irlba(t(cpm.norm), n = 5)
      r2 <- sapply(
          1:2, 
          function(k){summary(lm(ls ~ pca$x[, 1:k]))$r.squared
      })
      ls.pca <- mean(r2)
      corr.genes.ls <- correls(
        y = ls,
        x = t(cpm.norm),
        type = "spearman"
        )
      corr.genes.ls <- as.data.frame(corr.genes.ls)
      corr.genes.ls.corr <- mean(abs(corr.genes.ls$correlation) > 0.3)
      gene.corr.median <- median(
          abs(corr.genes.ls$correlation[abs(corr.genes.ls$correlation) > 0.3]), 
          na.rm = TRUE
          )
      corr.genes.ls$adj.pval <- p.adjust(
          corr.genes.ls$`p-value`, 
          method = "BH"
          )
      corr.genes.ls.pavl <- mean(corr.genes.ls$adj.pval < 0.05 )
      corr.genes.ls.pavl.corr <- mean(
          corr.genes.ls$`p-value` <0.05 & abs(corr.genes.ls$correlation) > 0.3
          )
      return(data.frame(
        file = file,
        pca = ls.pca,
        gene.corr.mde = gene.corr.median,
        gene.corr = corr.genes.ls.corr,
        gene.pval = corr.genes.ls.pavl,
        gene.pval.corr = corr.genes.ls.pavl.corr
      ))

    }, error = function(e) {
      message("Error in ", file, ": ", e$message)
      return(data.frame(
        file = file,
        pca = NA,
        gene.corr.mde = NA,
        gene.corr = NA,
        gene.pval = NA,
        gene.pval.corr = NA
      ))
    }),
    warning = function(w) {
      message("Warning in ", file, ": ", w$message)
      invokeRestart("muffleWarning")
    }
  )
}
all.genes.pca.ls <- pbmclapply(
  all.studies,
  genes.pca.ls,
  mc.cores = 15, 
  mc.preschedule = FALSE
  )
all.genes.pca.ls <- as.data.frame(do.call(rbind, all.genes.pca.ls))
```

### Batch Effects on Gene Expression and PCA

To assess the extent of batch effects across studies, sample-level batch
structure is estimated using model-based clustering (Gaussian mixture models
via `mclust`) applied to the median of the relative log expression (RLE)
values on CPM data. RLE medians serve as a proxy for sample-level technical variation: a
large spread of RLE medians within a study indicates the presence of
substantial unwanted variation. Two complementary metrics are then computed:
(1) the proportion of genes showing significant differential expression across
estimated batches (one-way ANOVA, Benjamini-Hochberg adjusted $p < 0.05$),
and (2) the association between batch membership and the first two principal
components, quantified using vector (canonical) correlation. Studies are
processed in chunks of 100 to manage memory usage.

```{r}
genes.pca.batch <- function(file) {
  withCallingHandlers(
    expr = tryCatch({
      temp.data <- qs::qread(file)
      gene.info <- rowData(temp.data)
      raw.data <- assay(temp.data[gene.info$gene_type == 'protein_coding', ], "raw_counts")
      keep.genes <- edgeR::filterByExpr(y = raw.data)
      raw.data <- raw.data[keep.genes, , drop = FALSE]
      ls <- log2(colSums(raw.data))
      q1 <- quantile(ls, 0.25)
      q3 <- quantile(ls, 0.75)
      iqr <- q3 - q1
      lower <- q1 - 1.5 * iqr
      upper <- q3 + 1.5 * iqr
      outliers <- !ls < lower | ls > upper
      raw.data <- raw.data[, outliers]
      ls <- ls[outliers]
      cpm.norm <- edgeR::cpm(raw.data, log = TRUE)
      rle.data <- cpm.norm - rowMedians(cpm.norm)
      rle.med <- colMedians(rle.data)
      batches <- mclust::Mclust(data = rle.med)
      anova.genes.batches <- row_oneway_equalvar(
          x = cpm.norm, 
          g = batches$classification
          )
      anova.genes.batches$pvalue <- p.adjust(
          p = anova.genes.batches$pvalue, 
          method = "BH"
          )
      gene.batches <- sum(anova.genes.batches$pvalue < 0.05)
      gene.batches <- gene.batches/nrow(cpm.norm)
      pca <- irlba::prcomp_irlba(t(cpm.norm), n = 5)
      r3 <- sapply(1:2, function(k){
        summary(lm(rle.med ~ pca$x[, 1:k]))$r.squared
      })
      rle.pca <- mean(r3)
      corr.genes.rle.med <- correls(
          y = rle.med,
          x = t(cpm.norm),
          type = "spearman"
          )
      corr.genes.rle.med <- as.data.frame(corr.genes.rle.med)
      corr.genes.rle.med <- sum(abs(corr.genes.rle.med$correlation) > 0.3 ) / nrow(cpm.norm)
      catvar.dummies <- fastDummies::dummy_cols(batches$classification)
      catvar.dummies <- catvar.dummies[, c(2:ncol(catvar.dummies))]
      vector.corr <- sapply(1:2, function(z) {
          cca <- cancor(x = pca$x[, 1:z, drop = FALSE], y = catvar.dummies)
          1 - prod(1 - cca$cor^2)
      })
      vector.corr <- mean(vector.corr)
      return(data.frame(
        file = file,
        rle.pca = rle.pca,
        gene.rle = corr.genes.rle.med,
        batches = length(unique(batches$classification)),
        gene.batches = gene.batches,
        vector.corr = vector.corr
      ))

    }, error = function(e) {
      message("Error in ", file, ": ", e$message)
      return(data.frame(
        file = file,
        rle.pca = NA,
        gene.rle = NA,
        batches = NA,
        gene.batches = NA,
        vector.corr = NA
      ))
    }),
    warning = function(w) {
      message("Warning in ", file, ": ", w$message)
      invokeRestart("muffleWarning")
    }
  )
}
chunk.size <- 100
idx.list <- split(
    seq_along(all.studies), 
    ceiling(seq_along(all.studies) / chunk.size)
    )
results.list <- lapply(
    idx.list, 
    function(idx) {
        message("Running chunk: ", idx[1], " to ", idx[length(idx)])
        out <- pbmclapply(
            all.studies[idx], 
            genes.pca.batch, 
            mc.cores = 10, 
            mc.preschedule = FALSE
            )
        do.call(rbind, out)
})
results.list <- do.call(rbind, results.list)
```

### Visualisation

The following plots summarise the gene-level and global measures of unwanted
variation across all 2,284 studies. Each point represents one study.

#### Library Size Effects

This scatter plot compares the global measure of library size association
($R^2$ from the first two PCs) against the gene-level measure (proportion of
genes with $|\rho| > 0.3$). A strong concordance between the two metrics
would indicate that global PCA-based summaries faithfully capture gene-level
library size effects.

```{r}
ggplot(all.genes.pca.ls, aes(x = pca, y = gene.corr )) +
    geom_point(alpha = 0.7, size = 2, color = 'khaki4') +
    xlab(expression(R^2)) +
    ylab("Number of genes (proportion)") +
    geom_abline(intercept = 0, slope = 1) +
    geom_smooth() +
    theme(
        panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 1),
        axis.title = element_text(size = 16),
        axis.text  = element_text(size = 14),
        legend.title = element_blank(),
        legend.text  = element_text(size = 14),
        legend.position = "none"
        )
all.genes.pca.ls.new <- all.genes.pca.ls
all.genes.pca.ls.new <- all.genes.pca.ls.new[rowSums(is.na(all.genes.pca.ls.new)) == 0 , ]
index <- all.genes.pca.ls.new$pca < 0.1 & all.genes.pca.ls.new$gene.corr > 0.05
sum(index) == 618
all.genes.pca.ls.new <- all.genes.pca.ls.new[ index, ]
range(all.genes.pca.ls.new$gene.corr) # 0.05072546 0.66689040
range(all.genes.pca.ls.new$gene.corr.mde) #  0.3213634 0.5943372

```

#### Batch Effects

This scatter plot compares the global measure of batch association (vector
correlation between batch membership and the first two PCs) against the
gene-level measure (proportion of genes significantly associated with
estimated batches). Studies in the upper-right quadrant exhibit strong batch
effects at both the gene and global levels.

```{r}
ggplot(batch.pca, aes(x = vector.corr, y = gene.batches)) +
    geom_point(alpha = 0.7, size = 2, color = 'khaki4') +
    xlab('Vector correlation') +
    ylab("Number of genes (proportion)") +
    geom_abline(intercept = 0, slope = 1) +
    geom_smooth() +
    theme(
        panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 1),
        axis.title = element_text(size = 16),
        axis.text  = element_text(size = 14),
        legend.title = element_blank(),
        legend.text  = element_text(size = 14),
        legend.position = "none"
    )
```

#### Gene Correlation with RLE Medians

This scatter plot compares the global $R^2$ of RLE medians explained by the
first two PCs against the proportion of genes correlated with RLE medians
($|\rho| > 0.3$). Studies with high values on both axes are those where
sample-level technical variation dominates the leading axes of transcriptome
variation.

```{r}
ggplot(batch.pca, aes(x = rle.pca, y = gene.rle)) +
    geom_point(alpha = 0.7, size = 2, color = 'khaki4') +
    xlab(expression(R^2)) +
    ylab("Number of genes (proportion)") +
    geom_abline(intercept = 0, slope = 1) +
    geom_smooth() +
    theme(
        panel.background = element_blank(),
        panel.grid = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 1),
        axis.title = element_text(size = 16),
        axis.text  = element_text(size = 14),
        legend.title = element_blank(),
        legend.text  = element_text(size = 14),
        legend.position = "none"
    )
```

### Number of Estimated Batches

This bar chart shows the distribution of the number of batches estimated by
`mclust` across all studies. The majority of studies are estimated to contain
between two and four batches, highlighting that unaccounted technical
heterogeneity is pervasive across publicly available RNA-seq data.

```{r}
df <- as.data.frame(table(batch.pca$batches))
df$vec <- as.numeric(as.character(df$Var1))
ggplot(df, aes(x = vec, y = Freq)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  scale_x_continuous(breaks = 1:9) + 
  labs(x = "Estimated batches", y = "Number of studies") +
  theme_minimal() +
  theme(
      panel.background = element_blank(),
      panel.grid = element_blank(),
      axis.line = element_line(colour = "black", linewidth = 1),
      axis.title = element_text(size = 16),
      axis.text  = element_text(size = 14),
      legend.title = element_blank(),
      legend.text  = element_text(size = 14),
      legend.position = "none"
  )
```
