---
title: "GTEx"
output: html_document
date: "2025-10-17"
editor_options: 
  chunk_output_type: console
---

# Libraries
```{r}
library(SummarizedExperiment)
library(matrixStats)
library(data.table)
library(patchwork)
library(parallel)
library(RUVprps)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(dplyr)
library(edgeR)
```

# Reading the data
```{r}
gtex.expr             <- fread('GTEx_Analysis_2022-06-06_v10_RNASeQCv2.4.2_gene_reads.gct')
gene.annot            <- gtex.expr[, c(1:2)]
gene.annot$gene.names <- gene.annot$Description
index                 <- duplicated(gene.annot$Description)
gene.annot$gene.names[index] <- paste0(
    gene.annot$Description[index], 
    '|', 
    gene.annot$Name[index]
    )
gtex.expr            <- gtex.expr[ ,-(1:2) ]
gtex.expr            <- as.matrix(gtex.expr)
row.names(gtex.expr) <- gene.annot$gene.names

sample.annot <- read.delim(
    file      = 'Gtex_SampleAnnot_V10.txt',
    row.names = 1
    )
common.samples <- intersect(
    colnames(gtex.expr), 
    row.names(sample.annot)
    ) # 19788
length(common.samples) == 19788
sample.annot <- sample.annot[common.samples , ]
all.equal(colnames(gtex.expr), row.names(sample.annot))
```

# Adding more possible batch
```{r}
sample.annot <- sample.annot[ , 1:19]
sample.annot$year <- unlist(lapply(
    sample.annot$SMNABTCHD,
    function(x) strsplit(x, '[/]')[[1]][3]))
sample.annot$year <- gsub('20', '', sample.annot$year)

sample.annot$month <- unlist(lapply(
    sample.annot$SMNABTCHD,
    function(x) strsplit(x, '[/]')[[1]][1]))

sample.annot$SMUBRID.id <- unlist(lapply(
    sample.annot$SMUBRID,
    function(x) strsplit(x, '[:]')[[1]][1]))
sample.annot$SMUBRID.no <- unlist(lapply(
    sample.annot$SMUBRID,
    function(x) strsplit(x, '[:]')[[1]][2]))

sample.annot$SMNABTCH.id <- unlist(lapply(
    sample.annot$SMNABTCH,
    function(x) strsplit(x, '[-]')[[1]][1]))

sample.annot$SMNABTCH.no <- unlist(lapply(
    sample.annot$SMNABTCH,
    function(x) strsplit(x, '[-]')[[1]][2]))

sample.annot$SMGEBTCH.id <- unlist(lapply(
    sample.annot$SMGEBTCH,
    function(x) strsplit(x, '[-]')[[1]][1]))
sample.annot$SMGEBTCH.no <- unlist(lapply(
    sample.annot$SMGEBTCH,
    function(x) strsplit(x, '[-]')[[1]][2]))

sample.annot$sample.a <- unlist(lapply(
    row.names(sample.annot),
    function(x) strsplit(x, '[-]')[[1]][1]))
sample.annot$sample.b <- unlist(lapply(
    row.names(sample.annot),
    function(x) strsplit(x, '[-]')[[1]][2]))
sample.annot$sample.c <- unlist(lapply(
    row.names(sample.annot),
    function(x) strsplit(x, '[-]')[[1]][3]))
sample.annot$sample.d <- unlist(lapply(
    row.names(sample.annot),
    function(x) strsplit(x, '[-]')[[1]][4]))
sample.annot$sample.e <- unlist(lapply(
    row.names(sample.annot),
    function(x) strsplit(x, '[-]')[[1]][5]))
sum(is.na(sample.annot$SMTSISCH))
index <- !is.na(sample.annot$SMTSISCH)
sample.annot <- sample.annot[index , ]
gtex.expr <- gtex.expr[ , index]
all.equal(colnames(gtex.expr), row.names(sample.annot))

library(dplyr)
sample.annot <- arrange(sample.annot, SMTSISCH)
gtex.expr <- gtex.expr[ , row.names(sample.annot)]
sample.annot$SMTSISCH.g <- cut(sample.annot$SMTSISCH, 6)
sample.annot$SMRIN <- cut(sample.annot$SMRIN, 6)
table(sample.annot$SMTSISCH.g, sample.annot$year)
```

# Creating RLE plots after CPM normalization
```{r}
tissues <- names(table(sample.annot$SMTSD)[table(sample.annot$SMTSD) > 20]) # 56
all.rle <- mclapply(tissues, function(x) {
    message("Processing: ", x)
    sub.info <- sample.annot[sample.annot$SMTSD == x , ]
    sub.data <- gtex.expr[, row.names(sub.info)]
    se.obj <- SummarizedExperiment(
        assays = list(RawCount = sub.data),
        colData = sub.info
        )
    tcga.se.obj <- prepareSeObj(
        data = se.obj,
        raw.count.assay.name = 'RawCount',
        remove.lowly.expressed.genes = TRUE,
        estimate.tumor.purity = 'both',
        assay.name.to.estimate.purity = 'RawCount',
        scale.singscore.values = FALSE,
        calculate.library.size = TRUE,
        add.immun.stroma.genes = FALSE,
        add.gene.details = FALSE,
        add.housekeeping.genes = FALSE,
        column.name = 'gene_name',
        gene.group = 'hgnc_symbol'
        )
    tcga.se.obj <- RUVprps:::orderSeObj(
        se.obj = tcga.se.obj,
        factors.to.order = c('SMTSISCH')
        )
    tcga.se.obj <- applyOtherNormalizations(
        se.obj = tcga.se.obj, 
        assay.name = 'RawCount', 
        new.assay.name = 'CPM'
        )
    tcga.se.obj <- computeRLE(se.obj = tcga.se.obj)
    rle.data <- data.frame(
        rle = tcga.se.obj@metadata$Metrics$CPM$global.level$RLE$gene.median.center$general$data$rle.med,
        sample = 1:ncol(tcga.se.obj),
        tissue = x
    )
    return(rle.data)
}, mc.cores = 14)

all.rle <- do.call(rbind, all.rle)
pdf(paste0(gtex.fig, 'RUVprps_GTEx_AllData_RleAfterCpmNorm.pdf'), width = 16, height = 12)
ggplot(all.rle, aes(x = sample, y = rle)) +
    geom_point(size = .4) +
    geom_smooth() +
    xlab('Samples') +
    ylab('RLE medians') +
    facet_wrap(~tissue, scales = "free", nrow = 10)
dev.off()
```

# Assessing all variation
## Immune genes-RLE approach
```{r}
load('kh.im.genes.rda')
head(kh.im.genes)
norm.hk <- kh.im.genes$hgnc_symbol[kh.im.genes$single.cell.rnaseq.hk.genes == 'yes']
pp.hk <- kh.im.genes$hgnc_symbol[kh.im.genes$immune.gene.signature == 'yes']
all.hk <- c(norm.hk, pp.hk)

tissues <- names(table(sample.annot$SMTSD)[table(sample.annot$SMTSD) > 20])
all.esti.immune <- mclapply(
    tissues,
    function(x) {
        message("Processing: ", x)
        sub.info <- sample.annot[sample.annot$SMTSD == x , ]
        sub.data <- gtex.expr[, row.names(sub.info)]
        se.obj <- SummarizedExperiment(assays = list(RawCount = sub.data),
                                       colData = sub.info)
        tcga.se.obj <- prepareSeObj(
            data = se.obj,
            raw.count.assay.name = 'RawCount',
            remove.lowly.expressed.genes = TRUE,
            estimate.tumor.purity = 'both',
            assay.name.to.estimate.purity = 'RawCount',
            scale.singscore.values = FALSE,
            calculate.library.size = TRUE,
            add.immun.stroma.genes = FALSE,
            add.gene.details = FALSE,
            add.housekeeping.genes = FALSE,
            column.name = 'gene_name',
            gene.group = 'hgnc_symbol'
            )
        ncg <- intersect(all.hk, row.names(tcga.se.obj))
        tcga.se.obj <- applyOtherNormalizations(
            se.obj = tcga.se.obj,
            assay.name = 'RawCount',
            new.assay.name = 'cpm',
            apply.log = TRUE
            )
        temp.data <- assay(x = tcga.se.obj, 'cpm')
        temp.data <- singscore::rankGenes(temp.data)
        immune.scores <- singscore::simpleScore(
            rankData = temp.data, 
            upSet = pp.hk
            )
        immune.scores <- cut(immune.scores$TotalScore, breaks = 3)
        tcga.se.obj$immune.scores <- immune.scores
        tcga.se.obj <- orderSeObj(
            se.obj = tcga.se.obj, 
            factors.to.order = 'immune.scores'
            )
        tcga.se.obj <- tryCatch({
            identifyUnknownUV(
                se.obj = tcga.se.obj,
                assay.name = 'cpm',
                approach = 'rle',
                apply.log = FALSE,
                ncg = NULL ,
                chronological.detection = TRUE,
                clustering.methods = 'mclust',
                assess.bio.association = FALSE,
                assess.uv.association = FALSE,
                generate.association.plot = FALSE,
                add.to.sample.annotation = TRUE,
                col.name = 'Estimated.batches',
                output.name = 'iteration1'
            )
        }, error = function(e) {
            message("identifyUnknownUV failed for ", x, ": ", e$message)
            return(NULL)
        })
        # If identifyUnknownUV() failed, skip association part
        if (is.null(tcga.se.obj) ||
            !"Estimated.batches" %in% colnames(colData(tcga.se.obj))) {
            return(data.frame(
                pval = NA,
                sig.test = NA,
                batches = NA
            ))
        }
        conf.mat <- data.frame(
            batch = paste0(tcga.se.obj$immune.scores),
            estimate.batch = tcga.se.obj$Estimated.batches
            )
        conf.mat <- table(conf.mat$batch, conf.mat$estimate.batch)
        sig.test <- tryCatch({
            DescTools::CramerV(conf.mat, bias.correct = TRUE)
        }, error = function(e)
            NA)
        chi <- tryCatch({
            chisq.test(conf.mat)
        }, error = function(e)
            list(p.value = NA))

        return(data.frame(
            pval = chi$p.value,
            sig.test = sig.test,
            batches = length(unique(tcga.se.obj$Estimated.batches))
        ))
    }, mc.cores = 14
    )

all.esti.immune <- do.call(rbind, all.esti.immune)
all.esti.immune$cancer <- tissues
all.esti.immune$type <- 'immune'
```

## Immune genes with RLE ncg approach
```{r}
all.esti.immune.ncg <- mclapply(
    tissues,
    function(x) {
        message("Processing: ", x)
        sub.info <- sample.annot[sample.annot$SMTSD == x , ]
        sub.data <- gtex.expr[, row.names(sub.info)]
        gtex.se.obj <- SummarizedExperiment(
            assays = list(RawCount = sub.data),
            colData = sub.info
            )
        gtex.se.obj <- prepareSeObj(
            data = gtex.se.obj,
            raw.count.assay.name = 'RawCount',
            remove.lowly.expressed.genes = TRUE,
            estimate.tumor.purity = NULL,
            assay.name.to.estimate.purity = 'RawCount',
            scale.singscore.values = FALSE,
            calculate.library.size = TRUE,
            add.immun.stroma.genes = FALSE,
            add.gene.details = FALSE,
            add.housekeeping.genes = FALSE,
            column.name = 'gene_name',
            gene.group = 'hgnc_symbol'
            )
        gtex.se.obj <- applyOtherNormalizations(
            se.obj = gtex.se.obj,
            assay.name = 'RawCount',
            new.assay.name = 'cpm',
            apply.log = TRUE
            )
        temp.data <- assay(x = gtex.se.obj, 'cpm')
        ranked.data <- singscore::rankGenes(temp.data)
        immune.scores <- singscore::simpleScore(
            rankData = ranked.data, 
            upSet = pp.hk
            )
        immune.scores <- cut(immune.scores$TotalScore, breaks = 3)
        gtex.se.obj$immune.scores <- immune.scores
        gtex.se.obj <- orderSeObj(
            se.obj = gtex.se.obj, 
            factors.to.order = 'immune.scores'
            )
        gtex.se.obj <- tryCatch({
            identifyUnknownUV(
                se.obj = gtex.se.obj,
                assay.name = 'cpm',
                approach = 'rle',
                apply.log = FALSE,
                ncg = intersect(row.names(gtex.se.obj), pp.hk) ,
                chronological.detection = TRUE,
                clustering.methods = 'mclust',
                assess.bio.association = FALSE,
                assess.uv.association = FALSE,
                generate.association.plot = FALSE,
                add.to.sample.annotation = TRUE,
                col.name = 'Estimated.batches',
                output.name = 'iteration1'
            )
        }, error = function(e) {
            message("identifyUnknownUV failed for ", x, ": ", e$message)
            return(NULL)
        })
        # If identifyUnknownUV() failed, skip association part
        if (is.null(gtex.se.obj) ||
            !"Estimated.batches" %in% colnames(colData(gtex.se.obj))) {
            return(data.frame(
                pval = NA,
                sig.test = NA,
                batches = NA
            ))
        }
        conf.mat <- data.frame(
            batch = paste0(gtex.se.obj$immune.scores),
            estimate.batch = gtex.se.obj$Estimated.batches
            )
        conf.mat <- table(conf.mat$batch, conf.mat$estimate.batch)
        sig.test <- tryCatch({
            DescTools::CramerV(conf.mat, bias.correct = TRUE)
        }, error = function(e)
            NA)
        
        chi <- tryCatch({
            chisq.test(conf.mat)
        }, error = function(e)
            list(p.value = NA))

        return(data.frame(
            pval = chi$p.value,
            sig.test = sig.test,
            batches = length(unique(gtex.se.obj$Estimated.batches))
        ))
    }, mc.cores = 14
    )

all.esti.immune.ncg <- do.call(rbind, all.esti.immune.ncg)
all.esti.immune.ncg$cancer <- tissues
all.esti.immune.ncg$type <- 'immune.rle.ncg'
```

## Smrin factor
```{r}
all.esti.smrin <- mclapply(
    tissues, 
    function(x) {
    message("Processing: ", x)
    sub.info <- sample.annot[sample.annot$SMTSD == x , ]
    sub.data <- gtex.expr[, row.names(sub.info)]
    gtex.se.obj <- SummarizedExperiment(
        assays = list(RawCount = sub.data),
        colData = sub.info
        )
    gtex.se.obj <- prepareSeObj(
        data = gtex.se.obj,
        raw.count.assay.name = 'RawCount',
        remove.lowly.expressed.genes = TRUE,
        estimate.tumor.purity = 'both',
        assay.name.to.estimate.purity = 'RawCount',
        scale.singscore.values = FALSE,
        calculate.library.size = TRUE,
        add.immun.stroma.genes = FALSE,
        add.gene.details = FALSE,
        add.housekeeping.genes = FALSE,
        column.name = 'gene_name',
        gene.group = 'hgnc_symbol'
        )
    gtex.se.obj <- applyOtherNormalizations(
        se.obj = gtex.se.obj,
        assay.name = 'RawCount',
        new.assay.name = 'cpm',
        apply.log = TRUE
        )
    gtex.se.obj <- orderSeObj(
        se.obj = gtex.se.obj, 
        factors.to.order = c('SMRIN')
        )
    gtex.se.obj <- tryCatch({
        identifyUnknownUV(
            se.obj = gtex.se.obj,
            assay.name = 'cpm',
            approach = 'rle',
            ncg = NULL,
            apply.log = FALSE,
            chronological.detection = TRUE,
            clustering.methods = 'mclust',
            assess.bio.association = FALSE,
            assess.uv.association = FALSE,
            generate.association.plot = FALSE,
            add.to.sample.annotation = TRUE,
            col.name = 'Estimated.batches',
            output.name = 'iteration1'
        )
    }, error = function(e) {
        message("identifyUnknownUV failed for ", x, ": ", e$message)
        return(NULL)
    })
    # If identifyUnknownUV() failed, skip association part
    if (is.null(gtex.se.obj) ||
        !"Estimated.batches" %in% colnames(colData(gtex.se.obj))) {
        return(data.frame(
            pval = NA,
            sig.test = NA,
            batches = NA
        ))
    }
    conf.mat <- data.frame(
        batch = droplevels(gtex.se.obj$SMRIN),
        estimate.batch = gtex.se.obj$Estimated.batches
        )
    conf.mat <- table(conf.mat$batch, conf.mat$estimate.batch)
    sig.test <- tryCatch({
        DescTools::CramerV(conf.mat, bias.correct = TRUE)
    }, error = function(e)
        NA)

    chi <- tryCatch({
        chisq.test(tbl)
    }, error = function(e)
        list(p.value = NA))
    
    return(data.frame(
        pval = chi$p.value,
        sig.test = sig.test,
        batches = length(unique(gtex.se.obj$Estimated.batches))
    ))
}, mc.cores = 14)

all.esti.smrin <- do.call(rbind, all.esti.smrin)
all.esti.smrin$cancer <- tissues
all.esti.smrin$type <- 'smrin'
```

## Smtsisch factor
```{r}
all.esti.smtsisch <- mclapply(tissues, function(x) {
    message("Processing: ", x)
    sub.info <- sample.annot[sample.annot$SMTSD == x , ]
    sub.data <- gtex.expr[, row.names(sub.info)]
    gtex.se.obj <- SummarizedExperiment(
        assays = list(RawCount = sub.data),
        colData = sub.info
        )
    gtex.se.obj <- prepareSeObj(
        data = gtex.se.obj,
        raw.count.assay.name = 'RawCount',
        remove.lowly.expressed.genes = TRUE,
        estimate.tumor.purity = NULL,
        assay.name.to.estimate.purity = 'RawCount',
        scale.singscore.values = FALSE,
        calculate.library.size = TRUE,
        add.immun.stroma.genes = FALSE,
        add.gene.details = FALSE,
        add.housekeeping.genes = FALSE,
        column.name = 'gene_name',
        gene.group = 'hgnc_symbol'
        )
    gtex.se.obj <- applyOtherNormalizations(
        se.obj = gtex.se.obj,
        assay.name = 'RawCount',
        new.assay.name = 'cpm',
        apply.log = TRUE
        )
    gtex.se.obj <- orderSeObj(
        se.obj = gtex.se.obj, 
        factors.to.order = c('SMTSISCH.g')
        )
    gtex.se.obj <- tryCatch({
        identifyUnknownUV(
            se.obj = gtex.se.obj,
            assay.name = 'cpm',
            approach = 'rle',
            nb.pcs = 3,
            ncg = NULL,
            apply.log = FALSE,
            chronological.detection = TRUE,
            clustering.methods = 'mclust',
            assess.bio.association = FALSE,
            assess.uv.association = FALSE,
            generate.association.plot = FALSE,
            add.to.sample.annotation = TRUE,
            col.name = 'Estimated.batches',
            output.name = 'iteration1'
        )
    }, error = function(e) {
        message("identifyUnknownUV failed for ", x, ": ", e$message)
        return(NULL)
    })
    # If identifyUnknownUV() failed, skip association part
    if (is.null(gtex.se.obj) ||
        !"Estimated.batches" %in% colnames(colData(gtex.se.obj))) {
        return(data.frame(
            pval = NA,
            sig.test = NA,
            batches = NA
        ))
    }
    conf.mat <- data.frame(
        batch = droplevels(gtex.se.obj$SMTSISCH.g),
        estimate.batch = gtex.se.obj$Estimated.batches
    )
    conf.mat <- table(conf.mat$batch, conf.mat$estimate.batch)
    sig.test <- tryCatch({
        DescTools::CramerV(conf.mat, bias.correct = FALSE)
    }, error = function(e)
        NA)
    
    chi <- tryCatch({
        chisq.test(conf.mat)
    }, error = function(e)
        list(p.value = NA))
    
    return(data.frame(
        pval = chi$p.value,
        sig.test = sig.test,
        batches = length(unique(gtex.se.obj$Estimated.batches))
    ))
}, mc.cores = 14)
all.esti.smtsisch <- do.call(rbind, all.esti.smtsisch)
all.esti.smtsisch$cancer <- tissues
all.esti.smtsisch$type <- 'smtsisch'
```

## Putting all together
```{r}
all.res <- rbind(
    all.esti.immune, 
    all.esti.immune.ncg, 
    all.esti.smrin, 
    all.esti.smtsisch
    )
all.res <- all.res[order(all.res$sig.test, decreasing = TRUE) , ]
all.res$cancer <- gsub(' - ', '-', all.res$cancer)

all.res$cancer <- paste0(
  toupper(substring(all.res$cancer, 1, 1)),
  tolower(substring(all.res$cancer, 2))
  )
all.res$cancer <- gsub("\\s*\\([^\\)]*\\)", "", all.res$cancer)
all.res$cancer <- factor(all.res$cancer, levels = unique(all.res$cancer))
all.res <- all.res[all.res$sig.test != 'Inf' , ]

## changing names
all.res$type[all.res$type == 'immune'] <- 'Immune scores-a'
all.res$type[all.res$type == 'smrin'] <- 'Smrin'
all.res$type[all.res$type == 'smtsisch'] <- 'Smtsisch'
all.res$type[all.res$type == 'immune.rle.ncg'] <- 'Immune scores-b'
all.res <- all.res[!is.na(all.res$type) , ]

library(ggplot2)
pdf(paste0(gtex.fig, 'RUVprps_GTEx_EstimateUnKnownBatch_RLEandPCA.pdf'), width = 20, height = 5.2)
ggplot(all.res, aes(x = cancer, y = sig.test, group = cancer, color = type)) +
    geom_line(color = 'gray', size= 2) + 
    geom_point(size = 5) +
    xlab('Tissue types') +
    ylab('Association') +
    scale_color_manual(values = c('darkgreen', 'orange', 'purple4', 'cyan'))+
    labs(size = "Batches", color = 'Batches') +
     theme(
        axis.line = element_line(colour = 'black', linewidth = 1),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        plot.title = element_text(size = 14),
        axis.text.x = element_text(size = 13, angle = 35, hjust = 1),
        axis.text.y = element_text(size = 15),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        legend.background = element_blank(),
        legend.key = element_blank()
        )
dev.off()
# qs::qsave(all.res, 'RUVprps_GTEx_EstimateUnKnownBatch_RLEandPCA_AllResForPlotting.qs')
```

# Normalization by RUVIII-PRPS
## Case example - Adipose Tissue
### RUV normalization
#### Preparing SE object
```{r}

library(dplyr)
sample.annot            <- arrange(sample.annot, SMTSISCH)
gtex.expr               <- gtex.expr[ , row.names(sample.annot)]
sample.annot$SMTSISCH.g <- cut(sample.annot$SMTSISCH, 6)
sample.annot$SMRIN      <- cut(sample.annot$SMRIN, 6)

selected.tissue <- sample.annot$SMTS == 'Adipose Tissue'
expr.data       <- gtex.expr[ , selected.tissue]
dim(expr.data) # 59033  1301
sample.info     <- sample.annot[selected.tissue , ]
data.list       <- list(raw.count = expr.data)
se.obj <- prepareSeObj(
    data                         = data.list, 
    sample.annotation            = sample.info, 
    raw.count.assay.name         = 'raw.count',
    remove.lowly.expressed.genes = TRUE,
    calculate.library.size       = TRUE, 
    estimate.tumor.purity        = NULL, 
    create.gene.annotation       = TRUE, 
    add.gene.details             = FALSE, 
    gene.group                   = 'hgnc_symbol',
    add.housekeeping.genes       = TRUE
    ) # 20724 1301
se.obj$library.size <- log2(se.obj$library.size)
```

#### Selection of NCG
```{r}
se.obj <- renameVariables(
    se.obj        = se.obj, 
    current.names = c('SMTSD', 'SMTSISCH.g', 'library.size'), 
    new.names     = c('Smtsd', 'Smtsisch', 'Library.size')
    )
ncg.options <- c(
    'AnovaCorr.AcrossAllSamples', 
    'AnovaCorr.PerBatchPerBiology', 
    'LinearMixedModel', 
    'TwoWayAnova'
    )
for(i in 1:4){
    se.obj <- findNcgSupervised(
        se.obj               = se.obj,
        assay.name           = 'raw.count',
        bio.variables        = c('Smtsd'),
        uv.variables         = c('Smtsisch', 'Library.size'), 
        samples.to.use       = 'all',
        approach             = ncg.options[i],
        form                 = ~ (1|Smtsd) + (1|Smtsisch) +  Library.size,
        use.rank             = FALSE, 
        bio.percentile       = NULL,
        uv.percentile        = 0.8, 
        pseudo.count         = 0.5, 
        adjust.data          = TRUE,
        adjustment.variables = "uv",
        ncg.selection.method = 'quantile',
        nb.ncg               = 0.05, 
        ncg.group.name       = 'getex',
        ncg.set.name         = ncg.options[i], 
        nb.cores             = 15,
        check.se.obj         = FALSE,
        assess.ncg           = TRUE)
}

countNCG(
    se.obj = se.obj, 
    ncg.selection = 'supervised', 
    ncg.group.name = 'getex', 
    create.venn.diagram = FALSE
    )
compareNCGs(
    se.obj = se.obj, 
    assay.name = 'raw.count',
    bio.variables = 'Smtsd',
    uv.variables = c('Smtsisch', 'Library.size'), 
    ncg.type = 'supervised', 
    ncg.group.name = 'getex'
    )
```

#### Creating PRPS
```{r}
se.obj@metadata$PRPS <- NULL
se.obj <- createPrPsSupervised(
    se.obj                   = se.obj, 
    assay.name               = 'raw.count', 
    bio.variables            = c('Smtsd'), 
    samples.to.use           = 'all',
    uv.variables             = c('Smtsisch', 'Library.size'),
    bio.clustering.method    = "kmeans",
    nb.bio.clusters          = 4, 
    apply.log                = FALSE,
    check.prps.connectedness = FALSE,
    apply.other.uv.variables = TRUE,
    prps.group.name          = 'getex',
    verbose                  = TRUE
    )

ggarrange(
    se.obj@metadata$PRPS$supervised$getex$`Smtsisch|Smtsd|Library.size|raw.count`$prps.map.plot, 
    se.obj@metadata$PRPS$supervised$getex$`Library.size|Smtsd|Smtsisch|raw.count`$prps.map.plot,
    common.legend = FALSE, 
    legend        = "right"
)
```

#### RUV-III normalization
```{r}
se.obj <- RUVIIIprps(
    se.obj               = se.obj, 
    assay.name           = 'raw.count', 
    control.sample.types = 'prps',
    prps.type            = 'supervised', 
    prps.group           = 'getex',
    prps.set.names       = 'all',
    ncg.type             = 'supervised', 
    ncg.group.names      = 'getex',
    ncg.set.names        = 'LinearMixedModel',
    k                    = c(1:16), 
    apply.log            = TRUE,
    data.to.log          = 'both', 
    pseudo.count         = 0.5,
    return.wa            = TRUE, 
    check.se.obj         = FALSE
    )
assessW(
    se.obj = se.obj, 
    bio.variables = c('Smtsd'), 
    uv.variables = c('Smtsisch', 'Library.size'), 
    compare.w = TRUE
    )
```

#### Assesss normalization
##### Gene variance composition
```{r}
se.obj <- createLogAssays(
    se.obj         = se.obj, 
    assay.names    = 'raw.count' , 
    replace.assays = TRUE, 
    new.name       = 'Log'
    )
names(assays(se.obj))[11] <- 'Raw.count'
se.obj <- computeGenesVarianceComposition(
    se.obj      = se.obj, 
    assay.names = c('Raw.count' , 'RUVIIIprps_k_10'), 
    form        = ~(1|Smtsd) + (1|Smtsisch) + (1|SMRIN) + Library.size,
    apply.log   = FALSE
    )
raw.gene.var <- se.obj@metadata$Metrics$Raw.count$gene.level$GeneVarDecompostion$LMM$`~Smtsd + Smtsisch + SMRIN + Library.size`$gene.variance
ruv.gene.var <- se.obj@metadata$Metrics$RUVIIIprps_k_10$gene.level$GeneVarDecompostion$LMM$`~Smtsd + Smtsisch + SMRIN + Library.size`$gene.variance

raw.gene.var %>% 
    data.frame() %>% 
    pivot_longer(everything(), names_to = 'Variable', values_to = 'Variance')  %>% 
    mutate(Variable = if_else(Variable == "SMRIN", "Smrin", Variable)) %>% 
    mutate(Variable = factor(Variable, levels = c("Smrin","Smtsisch","Library.size" ,"Smtsd","Residuals" ))) %>% 
   ggplot(aes(x = Variable, y = Variance, fill = Variable)) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = "grey30",
    alpha = 0.8) +
    ylim(0,1) +
  stat_summary(
    fun = median,
    geom = "point",
    size = 2,
    color = "black") +
  stat_summary(
    fun.data = median_hilow,
    fun.args = list(conf.int = 0.5),
    geom = "errorbar",
    width = 0.15,
    color = "black") +
  scale_fill_manual(values = selectColors(nb.color = 1:5, group = "pan.selection.b")) +
  labs(
    x = "",
    y = "Variance explained (%)") +
  theme_classic(base_size = 14) +
      theme(
        panel.background = element_blank(),
        legend.position  = 'none',
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.y     = element_text(size = 14),
        legend.text      = element_text(size = 18),
        axis.text.x      = element_text(size = 12, angle = 90, hjust = 1),
        axis.text.y      = element_text(size = 12, angle = 90))

ruv.gene.var %>% 
    data.frame() %>% 
    pivot_longer(everything(), names_to = 'Variable', values_to = 'Variance')  %>% 
    mutate(Variable = if_else(Variable == "SMRIN", "Smrin", Variable)) %>% 
    mutate(Variable = factor(Variable, levels = c("Smrin","Smtsisch","Library.size" ,"Smtsd","Residuals" ))) %>% 
   ggplot(aes(x = Variable, y = Variance, fill = Variable)) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = "grey30",
    alpha = 0.8) +
    ylim(0,1) +
  stat_summary(
    fun = median,
    geom = "point",
    size = 2,
    color = "black") +
  stat_summary(
    fun.data = median_hilow,
    fun.args = list(conf.int = 0.5),
    geom = "errorbar",
    width = 0.15,
    color = "black") +
  scale_fill_manual(values = selectColors(nb.color = 1:5, group = "pan.selection.b")) +
  labs(
    x = "",
    y = "Variance explained (%)") +
  theme_classic(base_size = 14) +
      theme(
        panel.background = element_blank(),
        legend.position  = 'none',
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.y     = element_text(size = 14),
        legend.text      = element_text(size = 18),
        axis.text.x      = element_text(size = 12, angle = 90, hjust = 1),
        axis.text.y      = element_text(size = 12, angle = 90))


data.frame(raw.count = raw.gene.var$Smtsd, ruv = ruv.gene.var$Smtsd) %>% 
    ggplot(.,aes(x = ruv, y = raw.count)) +
    geom_point(size = .5, alpha = .5, color = 'grey30') +
       geom_abline(
        slope = 1,
        intercept = 0,
        linewidth = 1,
        linetype = "dashed",
        color = 'red'
    ) +
    xlab('RUVprps_k_10') +
    ylab('Raw.count') +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 14),
        axis.title.y     = element_text(size = 14),
        axis.text.x      = element_text(size = 10),
        axis.text.y      = element_text(size = 10)
    )
```

##### PCA plots
```{r}
# PCA plots
hag <- raw.gene.var$Smtsisch > .2
sum(hag) == 1437

se.obj.a <- computePCA(
    se.obj    = se.obj[hag, ], 
    apply.log = FALSE
    )
se.obj.a <- RUVprps::plotPCA(
    se.obj        = se.obj.a, 
    variable      = 'Smtsd',
     points.alpha = .8, 
    points.size   = 2, 
    stroke.size   = .01
    )

se.obj.a@metadata$Metrics$Raw.count$global.level$PCA$fast.svd$Smtsd$scatter.plot +
    se.obj.a@metadata$Metrics$RUVIIIprps_k_10$global.level$PCA$fast.svd$Smtsd$scatter.plot +
      plot_layout(guides = "collect", ncol = 2) 


se.obj.a <- RUVprps::plotPCA(
    se.obj = se.obj.a, 
    variable = 'Smtsisch', 
    color.palette = 'pan.selection.b', 
      points.alpha = .8, 
    points.size = 2, 
    stroke.size = .01
    )

se.obj.a@metadata$Metrics$Raw.count$global.level$PCA$fast.svd$Smtsisch$scatter.plot +
    se.obj.a@metadata$Metrics$RUVIIIprps_k_10$global.level$PCA$fast.svd$Smtsisch$scatter.plot +
      plot_layout(guides = "collect", ncol = 2) 
```

##### RLE plots
```{r}
se.obj <- computeRLE(
    se.obj    = se.obj, 
    apply.log = FALSE
    )
se.obj <- plotRLE(
    se.obj          = se.obj, 
    variable        = 'Smtsisch', 
    variable.colors = selectColors(nb.color = 1:4, group = 'pan.selection.b')
    )

ggarrange(
    se.obj@metadata$Metrics$Raw.count$global.level$RLE$gene.median.center$Smtsisch$colored + 
        coord_cartesian(ylim = c(-2, 2)),
    se.obj@metadata$Metrics$RUVIIIprps_k_10$global.level$RLE$gene.median.center$Smtsisch$colored + 
        coord_cartesian(ylim = c(-2, 2)), 
    common.legend = TRUE
    )

```

##### Gene level anova anmd correlation
```{r}
se.obj <- computeGenesVariableAnova(
    se.obj         = se.obj, 
    assay.names    = 'all', 
    variable       = 'Smtsisch', 
    apply.log      = FALSE, 
    plot.top.genes = FALSE
    )
raw.anova <- se.obj@metadata$Metrics$Raw.count$gene.level$ANOVA$aov$Smtsisch$fstatistics.pvalues
ruv.anova <- se.obj@metadata$Metrics$RUVIIIprps_k_10$gene.level$ANOVA$aov$Smtsisch$fstatistics.pvalues

data.frame(Raw.count = raw.anova$pvalue, RUVprps_k_10 = ruv.anova$pvalue) %>% 
    pivot_longer(everything(), names_to = 'method', values_to = 'pval') %>% 
    ggplot( aes(x = pval, fill = method)) +
    geom_histogram(bins = 20) +
    xlab('p-values') +
    ylab('Frequency') +
    labs(fill = "Datasets") +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 14),
        axis.title.y     = element_text(size = 14),
        axis.text.x      = element_text(size = 10),
        axis.text.y      = element_text(size = 10),
        legend.position  = c(0.7, 0.5)
        )

se.obj <- computeGenesVariableCorrelation(
    se.obj = se.obj, 
    assay.names = 'all', 
    variable = 'SMTSISCH', 
    apply.log = FALSE
    )

raw.corr <- se.obj@metadata$Metrics$Raw.count$gene.level$Correlation$spearman$SMTSISCH$correlations.pvalues
ruv.corr <- se.obj@metadata$Metrics$RUVIIIprps_k_10$gene.level$Correlation$spearman$SMTSISCH$correlations.pvalues

data.frame(Raw.count = raw.corr[,2], RUVprps_k_10 = ruv.corr[,2]) %>% 
    pivot_longer(everything(), names_to = 'method', values_to = 'pval') %>% 
    ggplot( aes(x = pval, fill = method)) +
    geom_histogram(bins = 20) +
    xlab('Correlations') +
    ylab('Frequency') +
    labs(fill = "Datasets") +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 14),
        axis.title.y     = element_text(size = 14),
        axis.text.x      = element_text(size = 10),
        axis.text.y      = element_text(size = 10),
        legend.position  = 'none'
        )


## an example
raw.corr <- raw.corr[order(raw.corr[,2], decreasing = T) , ]
head(raw.corr)
raw.data <- assay(se.obj, 'Raw.count')
ruv.data <- assay(se.obj, 'RUVIIIprps_k_10')

data.frame(
    Raw.count = raw.data['ACSL1' , ],
    RUVprps_k_10 = ruv.data['ACSL1' , ],
    var = se.obj$SMTSISCH) %>%
    pivot_longer(-var, names_to = 'method', values_to = 'expr') %>%
    ggplot(aes(x = var, y = expr)) +
    geom_point(color = 'grey30', size = .5) +
    facet_wrap( ~ method) +
    xlab('SMTSISCH') +
    ylab('Gene expression (ACSL1)') +
    geom_smooth(color = 'red') +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 14),
        axis.title.y     = element_text(size = 14),
        axis.text.x      = element_text(size = 10),
        axis.text.y      = element_text(size = 10),
        strip.text.x     = element_text(size = 14)
    )

data.frame(
    Raw.count = raw.data['CCSER1' , ],
    RUVprps_k_10 = ruv.data['CCSER1' , ],
    var = se.obj$SMTSISCH) %>%
    pivot_longer(-var, names_to = 'method', values_to = 'expr') %>%
    ggplot(aes(x = var, y = expr)) +
    geom_point(color = 'grey30', size = .5) +
    facet_wrap( ~ method) +
    xlab('SMTSISCH') +
    ylab('Gene expression (CCSER1)') +
    geom_smooth(color = 'red') +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 14),
        axis.title.y     = element_text(size = 14),
        axis.text.x      = element_text(size = 10),
        axis.text.y      = element_text(size = 10),
        strip.text.x     = element_text(size = 14)
    )
```
