---
title: "Assess CCA-based approach to create PRPS"
output: html_document
date: "2026-05-26"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(SummarizedExperiment)
library(matrixStats)
library(parallel)
library(RUVprps)
library(cluster)
library(mclust)
library(ggpubr)
library(irlba)
library(dplyr)
library(edgeR)
library(tidyr)
library(grid)
```

# CLLE
## Reading the data
```{r}
omics.profile <- read.csv('OmicsProfiles.csv')
omics.profile <- omics.profile[omics.profile$DataType == 'rna' , ]
length(unique(omics.profile$ModelID)) # 1699
omics.profile$years <- substr(
    x     = omics.profile$SequencingDate, 
    start = 1, 
    stop  = 4
    )
omics.profile$months <- substr(
    x     = omics.profile$SequencingDate, 
    start = 6, 
    stop  = 7
    )
omics.profile$days <- substr(
    x     = omics.profile$SequencingDate, 
    start = 9, 
    stop  = 10
    )
omics.profile <- dplyr::arrange(
    omics.profile, 
    Lineage, 
    CellFormat, 
    Stranded, years
    )

expr.data           <- as.data.frame(data.table::fread('OmicsExpressionRawReadCountHumanAllGenes.csv'))
cell.info           <- expr.data[, 1:6]
expr.data           <- expr.data[ , -c(1:6)]
expr.data           <- t(expr.data)
colnames(expr.data) <- cell.info$ModelID
all.equal(colnames(expr.data), omics.profile$ModelID)
expr.data           <- expr.data[ , omics.profile$ModelID]
all.equal(colnames(expr.data), omics.profile$ModelID)
```


# Assessing PRPS using CCA
### Reading the data
```{r}
cell.annot <- omics.profile
all.equal(colnames(expr.data), cell.annot$ModelID)

remov.cells <- !cell.annot$Lineage == "" # 1742
cell.annot  <- cell.annot[remov.cells , ]
expr.data   <- expr.data[ , remov.cells ]

remove.cells <- !duplicated(cell.annot$ModelID)
cell.annot   <- cell.annot[remove.cells , ]
expr.data    <- expr.data[ , remove.cells ]

remove.cells <- cell.annot$Lineage %in% RUVprps:::findRepeatingPatterns(
    vec = cell.annot$Lineage, 
    n.repeat = 10
    )
cell.annot            <- cell.annot[remove.cells , ]
expr.data             <- expr.data[ , remove.cells ]
all.equal(colnames(expr.data), cell.annot$ModelID)
row.names(cell.annot) <- cell.annot$ModelID
```

### Prepare SE object and find HVG
```{r}
raw.data <- list(RawCount = expr.data)
se.obj <- prepareSeObj(
    data                          = raw.data, 
    sample.annotation             = cell.annot, 
    raw.count.assay.name          = 'RawCount', 
    remove.lowly.expressed.genes  = TRUE, 
    biological.group              = 'Lineage', 
    count.cutoff                  = 20,
    calculate.library.size        = TRUE, 
    estimate.tumor.purity         = NULL, 
    assay.name.to.estimate.purity = 'RawCount'
    )
se.obj$library.size <- log2(se.obj$library.size)
se.obj <- findHVG(
    se.obj     = se.obj, 
    assay.name = 'RawCount', 
    nb.cores   = 14
    )
hvg <- se.obj@metadata$HVG$hvg_vst$`3638|acrossAllSamples|RawCount`$hvg.set
sum(hvg) == 3638
```

### Plotting uMAP raw data
```{r}
set.seed(2234)
pca.all <- irlba::prcomp_irlba(
    x = t(log2(expr.data+1)), 
    n = 20
    )
set.seed(2234)
umap.all <- uwot::umap(
    pca.all$x, 
    n_neighbors = 30, 
    min_dist    = 0.3, 
    metric      = "cosine"
    )
# plotting
data.frame(
  umap1                     = umap.all[, 1],
  umap2                     = umap.all[, 2],
  group                     = cell.annot$Lineage) %>%
  ggplot(aes(x              = umap1, y = umap2, color = group)) +
  geom_point(alpha          = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color                =1:34,
    group                   = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10),
    legend.position  = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points


data.frame(
  umap1            = umap.all[, 1],
  umap2            = umap.all[, 2],
  group            = cell.annot$Stranded) %>%
  ggplot(aes(x     = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c('darkgreen', 'cyan')) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10),
    legend.position  = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points
```

### Finding PRPS using CCA
#### Using 40 CCA
##### PRPS
```{r}
set.seed(4040)
se.obj <- createPrPsUnSupervisedByCca(
    se.obj             = se.obj, 
    assay.name         = 'RawCount', 
    main.uv.variable   = 'Stranded', 
    nb.cca             = 40, 
    hvg                = hvg,
    coordinates.to.use = 'cca', 
    max.prps.sets      = 50,
    )
mnn          <- se.obj@metadata$PRPS$un.supervised$Stranded$`Stranded|CcaPca|RawCount`$mnn.samples
dim(mnn) # 23 4
sample.annot <- as.data.frame(colData(se.obj))
a            <- sample.annot[mnn$left.samples, ]$Lineage
b            <- sample.annot[mnn$right.samples, ]$Lineage
tbl          <- table(a, b)

ComplexHeatmap::Heatmap(
    matrix = tbl , 
    name = 'Frequency',
    cluster_columns = F,
    cluster_rows = F,
    show_row_dend = F, 
    show_column_dend = F,
    column_names_rot = 90,
    col = c('white', 'tomato'),
    heatmap_legend_param = list(
    title = "Frequency",
    direction = "horizontal",
    title_gp = gpar(fontsize = 14, fontface = "bold"),  # legend title size
    labels_gp = gpar(fontsize = 14),                     # legend label size
    legend_height = unit(5, "cm"),                       # make legend taller
    legend_width = unit(1.2, "cm")                       # make legend wider (if horizontal)
    ))
```

##### Plotting umap using CCA
```{r}
set.seed(4040)
se.obj <- RUVprps:::computeCCA(
    se.obj     = se.obj, 
    assay.name = 'RawCount', 
    nb.cca     = 40, 
    variable   = 'Stranded', 
    hvg        = hvg
    )
cca.corr <- se.obj@metadata$CCA$RawCount$Stranded$`Cca:40|HVG:36380|Scale:TRUE|Norm:CPM`$False_True
cca.corr <- cca.corr[cell.annot$ModelID , ]
set.seed(4040)
umap.all <- uwot::umap(
    cca.corr, 
    n_neighbors = 30, 
    min_dist    = 0.3, 
    metric      = "cosine"
    )
# plotting
data.frame(
  umap1                     = umap.all[, 1],
  umap2                     = umap.all[, 2],
  group                     = cell.annot$Lineage) %>%
  ggplot(aes(x              = umap1, y = umap2, color = group)) +
  geom_point(alpha          = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color                =1:34,
    group                   = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10),
    legend.position  = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points


data.frame(
  umap1            = umap.all[, 1],
  umap2            = umap.all[, 2],
  group            = cell.annot$Stranded) %>%
  ggplot(aes(x     = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c('darkgreen', 'cyan')) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10),
    legend.position  = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points
```

##### Silhouette plot
```{r}
## silhouette
set.seed(2234)
pca.all           <- irlba::prcomp_irlba(x = t(log2(expr.data+1)), n = 40)
d.matrix          <- as.matrix(dist(pca.all$x[,1:40]))
avg.width.biology <- summary(silhouette(as.numeric(as.factor(se.obj$Lineage)), d.matrix))$avg.width
avg.width.strand  <- summary(silhouette(as.numeric(as.factor(se.obj$Stranded)), d.matrix))$avg.width

all.equal(row.names(cca.corr), cell.annot$ModelID)
d.matrix              <- as.matrix(dist(cca.corr[,1:40]))
avg.width.biology.cca <- summary(silhouette(as.numeric(as.factor(cell.annot$Lineage)), d.matrix))$avg.width
avg.width.strand.caa  <- summary(silhouette(as.numeric(as.factor(cell.annot$Stranded)), d.matrix))$avg.width

data.frame(
    'Lineage\n(PCA)'                     = avg.width.biology, 
    'Stranded\n(PCA)'                    = avg.width.strand, 
    'Lineage\n(CCA)'                     = avg.width.biology.cca,
    'Stranded\n(CCA)'                    = avg.width.strand.caa ,
    check.names                          = FALSE) %>% 
    pivot_longer(everything(), values_to = 'val', names_to = 'method') %>% 
    ggplot(., aes(x                      = method, y = val)) +
    ylab('Silhouette\ncoefficient') +
    xlab('') +
    geom_col() +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 16),
        axis.title.y     = element_text(size = 16),
        axis.text.x      = element_text(size = 12),
        axis.text.y      = element_text(size = 12),
        legend.title     = element_blank(),
        legend.text      = element_text(size = 10),
        legend.position  = 'none'
    )
```

#### Using 30 CCA
##### PRPS
```{r}
set.seed(3030)
se.obj <- createPrPsUnSupervisedByCca(
    se.obj             = se.obj, 
    assay.name         = 'RawCount', 
    main.uv.variable   = 'Stranded', 
    nb.cca             = 30, 
    hvg                = hvg,
    coordinates.to.use = 'cca', 
    max.prps.sets      = 50,
    )
mnn          <- se.obj@metadata$PRPS$un.supervised$Stranded$`Stranded|CcaPca|RawCount`$mnn.samples
dim(mnn) # 21 4
sample.annot <- as.data.frame(colData(se.obj))
a            <- sample.annot[mnn$left.samples, ]$Lineage
b            <- sample.annot[mnn$right.samples, ]$Lineage
tbl          <- table(a, b)

ComplexHeatmap::Heatmap(
    matrix               = tbl , 
    name                 = 'Frequency',
    cluster_columns      = F,
    cluster_rows         = F,
    show_row_dend        = F, 
    show_column_dend     = F,
    column_names_rot     = 90,
    col                  = c('white', 'tomato'),
    heatmap_legend_param = list(
    title                = "Frequency",
    direction            = "horizontal",
    title_gp             = gpar(fontsize = 14, fontface = "bold"),
    labels_gp            = gpar(fontsize = 14),
    legend_height        = unit(5, "cm"),
    legend_width         = unit(1.2, "cm") 
    ))
```

##### Plotting umap using CCA
```{r}
set.seed(3030)
se.obj <- RUVprps:::computeCCA(
    se.obj     = se.obj, 
    assay.name = 'RawCount', 
    nb.cca     = 30, 
    variable   = 'Stranded', 
    hvg        = hvg
    )
cca.corr <- se.obj@metadata$CCA$RawCount$Stranded$`Cca:30|HVG:36380|Scale:TRUE|Norm:CPM`$False_True
cca.corr <- cca.corr[cell.annot$ModelID , ]
set.seed(3030)
umap.all <- uwot::umap(
    cca.corr, 
    n_neighbors = 30, 
    min_dist = 0.3, 
    metric = "cosine"
    )
# plotting
data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = cell.annot$Lineage) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color =1:34,
    group = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points


data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = cell.annot$Stranded) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c('darkgreen', 'cyan')) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points
```

##### Silhouette plot
```{r}
## silhouette
d.matrix          <- as.matrix(dist(pca.all$x[,1:30]))
avg.width.biology <- summary(silhouette(as.numeric(as.factor(se.obj$Lineage)), d.matrix))$avg.width
avg.width.strand  <- summary(silhouette(as.numeric(as.factor(se.obj$Stranded)), d.matrix))$avg.width

all.equal(row.names(cca.corr), cell.annot$ModelID)
d.matrix              <- as.matrix(dist(cca.corr[,1:30]))
avg.width.biology.cca <- summary(silhouette(as.numeric(as.factor(cell.annot$Lineage)), d.matrix))$avg.width
avg.width.strand.caa  <- summary(silhouette(as.numeric(as.factor(cell.annot$Stranded)), d.matrix))$avg.width

data.frame(
    'Lineage\n(PCA)'                     = avg.width.biology, 
    'Stranded\n(PCA)'                    = avg.width.strand, 
    'Lineage\n(CCA)'                     = avg.width.biology.cca,
    'Stranded\n(CCA)'                    = avg.width.strand.caa ,
    check.names                          = FALSE) %>% 
    pivot_longer(everything(), values_to = 'val', names_to = 'method') %>% 
    ggplot(., aes(x                      = method, y = val)) +
    ylab('Silhouette\ncoefficient') +
    xlab('') +
    geom_col() +
    theme(
        panel.background = element_blank(),
        axis.line        = element_line(colour = 'black', linewidth = 1),
        axis.title.x     = element_text(size = 16),
        axis.title.y     = element_text(size = 16),
        axis.text.x      = element_text(size = 12),
        axis.text.y      = element_text(size = 12),
        legend.title     = element_blank(),
        legend.text      = element_text(size = 10),
        legend.position  = 'none'
    )
```

#### Using 20 CCA
##### PRPS
```{r}
set.seed(2020)
se.obj <- createPrPsUnSupervisedByCca(
    se.obj             = se.obj, 
    assay.name         = 'RawCount', 
    main.uv.variable   = 'Stranded', 
    nb.cca             = 20, 
    hvg                = hvg,
    coordinates.to.use = 'cca', 
    max.prps.sets      = 50,
    )
mnn          <- se.obj@metadata$PRPS$un.supervised$Stranded$`Stranded|CcaPca|RawCount`$mnn.samples
sample.annot <- as.data.frame(colData(se.obj))
a            <- sample.annot[mnn$left.samples, ]$Lineage
b            <- sample.annot[mnn$right.samples, ]$Lineage
tbl          <- table(a, b)

ComplexHeatmap::Heatmap(
    matrix               = tbl , 
    name                 = 'Frequency',
    cluster_columns      = F,
    cluster_rows         = F,
    show_row_dend        = F, 
    show_column_dend     = F,
    column_names_rot     = 90,
    col                  = c('white', 'tomato'),
    heatmap_legend_param = list(
    title                = "Frequency",
    direction            = "horizontal",
    title_gp             = gpar(fontsize = 14, fontface = "bold"),
    labels_gp            = gpar(fontsize = 14),
    legend_height        = unit(5, "cm"),
    legend_width         = unit(1.2, "cm")
    ))
```

##### Plotting umap using CCA
```{r}
set.seed(2020)
se.obj <- RUVprps:::computeCCA(
    se.obj     = se.obj, 
    assay.name = 'RawCount', 
    nb.cca     = 20, 
    variable   = 'Stranded', 
    hvg        = hvg
    )
cca.corr <- se.obj@metadata$CCA$RawCount$Stranded$`Cca:20|HVG:36380|Scale:TRUE|Norm:CPM`$False_True
cca.corr <- cca.corr[cell.annot$ModelID , ]
set.seed(2020)
umap.all <- uwot::umap(
    cca.corr, 
    n_neighbors = 30, 
    min_dist    = 0.3, 
    metric      = "cosine"
    )
# plotting
data.frame(
  umap1                     = umap.all[, 1],
  umap2                     = umap.all[, 2],
  group                     = cell.annot$Lineage) %>%
  ggplot(aes(x              = umap1, y = umap2, color = group)) +
  geom_point(alpha          = .4, size = 2, shape = 19, stroke = NA) +
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color                =1:34,
    group                   = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 10),
    legend.position  = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)

data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = cell.annot$Stranded) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c('darkgreen', 'cyan')) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = 'none'
  ) +
  guides(color = guide_legend(override.aes = list(size = 4)), ncol = 2)   # enlarge legend points
```

##### Silhouette plot
```{r}
## silhouette
d.matrix          <- as.matrix(dist(pca.all$x[,1:20]))
avg.width.biology <- summary(silhouette(as.numeric(as.factor(se.obj$Lineage)), d.matrix))$avg.width
avg.width.strand  <- summary(silhouette(as.numeric(as.factor(se.obj$Stranded)), d.matrix))$avg.width

all.equal(row.names(cca.corr), cell.annot$ModelID)
d.matrix              <- as.matrix(dist(cca.corr[,1:20]))
avg.width.biology.cca <- summary(silhouette(as.numeric(as.factor(cell.annot$Lineage)), d.matrix))$avg.width
avg.width.strand.caa  <- summary(silhouette(as.numeric(as.factor(cell.annot$Stranded)), d.matrix))$avg.width

data.frame(
    'Lineage\n(PCA)' = avg.width.biology, 
    'Stranded\n(PCA)' = avg.width.strand, 
    'Lineage\n(CCA)' = avg.width.biology.cca,
    'Stranded\n(CCA)' = avg.width.strand.caa ,
    check.names = FALSE) %>% 
    pivot_longer(everything(), values_to = 'val', names_to = 'method') %>% 
    ggplot(., aes(x = method, y = val)) +
    ylab('Silhouette\ncoefficient') +
    xlab('') +
    geom_col() +
    theme(
        panel.background = element_blank(),
        axis.line = element_line(colour = 'black', linewidth = 1),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.position = 'none'
    )
```



# Three large breast cancer
## Reading the data
```{r}
ms.se.obj <- qs::qread(
    file = 'ThreeLargeBreastCancer/Data/RUVprps_ThreeLargeBreastCancer_SEobject.qs'
    )
ms.se.obj <- ms.se.obj[ , !is.na(ms.se.obj$pam50)]
```

## UMAP plots of the all data
```{r}
expr.data <- assay(ms.se.obj, 'FPKM')
# dim: 13665 4573 
set.seed(2233)
pca.all <- irlba::prcomp_irlba(x = t(expr.data), n = 5)
set.seed(2233)
umap.all <- uwot::umap(
    pca.all$x, 
    n_neighbors = 30, 
    min_dist    = 0.3, 
    metric      = "cosine"
    )
# plotting
data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = ms.se.obj$pam50) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color =1:34,
    group = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 16),
    legend.position = 'bottom'
  ) +
  guides(color = guide_legend(override.aes = list(size = 6, alpha = 1), nrow = 1))   # enlarge legend points

data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = ms.se.obj$studies) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c("GSE81538" = 'darkgreen', "GSE96058" = 'tomato', "TCGA" = 'darkblue')) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 16),
    legend.position = 'bottom'
  ) +
  guides(color = guide_legend(override.aes = list(size = 6, alpha = 1), nrow = 1))    # enlarge legend points
```

## Overall PRPS
```{r}
ms.se.obj <- findHVG(
    se.obj        = ms.se.obj,
    assay.name    = 'FPKM',
    apply.log     = FALSE,
    approach      = 'lmm',
    form          = ~(1|pam50) + (1|studies),
    bio.variables = 'pam50',
    uv.variables  = 'studies',
    nb.hvg        = 0.05,
    nb.cores      = 14
    )
hvg <- ms.se.obj@metadata$HVG$hvg_lmm$`683|Bio:pam50|Uv:studies|FPKM`$hvg.set
sum(hvg) == 683
```

## PRPS for TCGA and GSE96058
### UMAP plots of CCs
```{r}
keep.samples <- ms.se.obj$studies %in% c('TCGA', 'GSE96058')
new.se.obj <- ms.se.obj[ , keep.samples]
set.seed(1122)
new.se.obj <- RUVprps:::computeCCA(
    se.obj        = new.se.obj, 
    assay.name    = 'FPKM', 
    apply.log     = FALSE,  
    normalization = NULL,
    nb.cca        = 30, 
    variable      = 'studies', 
    hvg           = hvg
    )
cca.corr <- new.se.obj@metadata$CCA$FPKM$studies$`Cca:30|HVG:13665|Scale:TRUE|Norm:`$TCGA_GSE96058
cca.corr <- cca.corr[colnames(new.se.obj) , ]

set.seed(1122)
umap.all <- uwot::umap(
    cca.corr, 
    n_neighbors = 30, 
    min_dist = 0.3, 
    metric = "cosine"
    )
# plotting
data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = new.se.obj$pam50) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color = 1:5,
    group = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 16)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)), ncol = 2)   # enlarge legend points

data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = new.se.obj$studies) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c("GSE81538" = 'darkgreen', "GSE96058" = 'tomato', "TCGA" = 'darkblue')) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 16)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)), ncol = 2)   # enlarge legend points
```

### PAM50 distibution
```{r}
tbl <- table(new.se.obj$studies, new.se.obj$pam50)
ht <- ComplexHeatmap::Heatmap(
    matrix = tbl , 
    name = 'Frequency',
    cluster_columns = F,
    cluster_rows = F,
    show_row_dend = F, 
    show_column_dend = F,
    column_names_rot = 45,
    col = c('white', 'tomato'),
    heatmap_legend_param = list(
    title = "Frequency",
    direction = "horizontal",
    title_gp = gpar(fontsize = 14, fontface = "bold"),
    labels_gp = gpar(fontsize = 14),
    legend_height = unit(5, "cm"),
    legend_width = unit(1.2, "cm")))
ComplexHeatmap::draw(ht, heatmap_legend_side = "bottom")
```

### Create PRPS and confusion matrix
```{r}
new.se.obj <- createPrPsUnSupervisedByCca(
    se.obj             = new.se.obj, 
    assay.name         = 'FPKM',
    main.uv.variable   = 'studies',
    hvg                = hvg, 
    coordinates.to.use = "cca",
    nb.cca             = 30, 
    nb.pcs             = 3,
    apply.log          = F, 
    min.nb.for.mnn     = 5,
    apply.log.for.prps = F, 
    normalization      = NULL,
    max.prps.sets      = 20
    )
mnn          <- new.se.obj@metadata$PRPS$un.supervised$studies$`studies|CcaPca|FPKM`$mnn.samples
dim(mnn)
sample.annot <- as.data.frame(colData(ms.se.obj))
a            <- sample.annot[mnn$left.samples, ]$pam50
b            <- sample.annot[mnn$right.samples, ]$pam50
tbl          <- table(a, b)

ComplexHeatmap::Heatmap(
    matrix               = tbl , 
    name                 = 'Frequency',
    cluster_columns      = F,
    cluster_rows         = F,
    show_row_dend        = F, 
    show_column_dend     = F,
    column_names_rot     = 90,
    col                  = c('white', 'tomato'),
    heatmap_legend_param = list(
    title                = "Frequency",
    direction            = "horizontal",
    title_gp             = gpar(fontsize = 14, fontface = "bold"),
    labels_gp            = gpar(fontsize = 14),
    legend_height        = unit(5, "cm"),
    legend_width         = unit(1.2, "cm")
    ))
```

## PRPS for TCGA and GSE81538
### UMAP plots of CCs
```{r}
keep.samples <- ms.se.obj$studies %in% c('TCGA', 'GSE81538')
new.se.obj <- ms.se.obj[ , keep.samples]
set.seed(1133)
new.se.obj <- RUVprps:::computeCCA(
    se.obj        = new.se.obj, 
    assay.name    = 'FPKM', 
    apply.log     = FALSE,  
    normalization = NULL,
    nb.cca        = 30, 
    variable      = 'studies', 
    hvg           = hvg
    )
cca.corr <- new.se.obj@metadata$CCA$FPKM$studies$`Cca:30|HVG:13665|Scale:TRUE|Norm:`$TCGA_GSE81538
cca.corr <- cca.corr[colnames(new.se.obj) , ]

set.seed(1133)
umap.all <- uwot::umap(
    cca.corr, 
    n_neighbors = 30, 
    min_dist    = 0.3, 
    metric      = "cosine"
    )
# plotting
data.frame(
  umap1            = umap.all[, 1],
  umap2            = umap.all[, 2],
  group            = new.se.obj$pam50) %>%
  ggplot(aes(x     = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color = 1:5,
    group = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 16)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)), ncol = 2)

data.frame(
  umap1            = umap.all[, 1],
  umap2            = umap.all[, 2],
  group            = new.se.obj$studies) %>%
  ggplot(aes(x     = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +
  xlab('UMAP1') +
  ylab('UMAP2') +
   scale_color_manual(values = c("GSE81538" = 'darkgreen', "GSE96058" = 'tomato', "TCGA" = 'darkblue')) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 16)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)), ncol = 2)   # enlarge legend points
```

### PAM50 distribution
```{r}
tbl <- table(new.se.obj$studies, new.se.obj$pam50)
ht <- ComplexHeatmap::Heatmap(
    matrix               = tbl , 
    name                 = 'Frequency',
    cluster_columns      = F,
    cluster_rows         = F,
    show_row_dend        = F, 
    show_column_dend     = F,
    column_names_rot     = 45,
    col                  = c('white', 'tomato'),
    heatmap_legend_param = list(
    title                = "Frequency",
    direction            = "horizontal",
    title_gp             = gpar(fontsize = 14, fontface = "bold"),
    labels_gp            = gpar(fontsize = 14),
    legend_height        = unit(5, "cm"),
    legend_width         = unit(1.2, "cm")))
ComplexHeatmap::draw(ht, heatmap_legend_side = "bottom")

```

### Create PRPS data
```{r}
new.se.obj <- createPrPsUnSupervisedByCca(
    se.obj             = new.se.obj, 
    assay.name         = 'FPKM',
    main.uv.variable   = 'studies',
    hvg                = hvg, 
    coordinates.to.use = "cca",
    nb.cca             = 30, 
    nb.pcs             = 3,
    apply.log          = F, 
    min.nb.for.mnn     = 5,
    apply.log.for.prps = F, 
    normalization      = NULL,
    max.prps.sets      = 20
    )
mnn          <- new.se.obj@metadata$PRPS$un.supervised$studies$`studies|CcaPca|FPKM`$mnn.samples
dim(mnn) # 50 4
sample.annot <- as.data.frame(colData(ms.se.obj))
a            <- sample.annot[mnn$left.samples, ]$pam50
b            <- sample.annot[mnn$right.samples, ]$pam50
tbl          <- table(a, b)

ComplexHeatmap::Heatmap(
    matrix = tbl , 
    name = 'Frequency',
    cluster_columns = F,
    cluster_rows = F,
    show_row_dend = F, 
    show_column_dend = F,
    column_names_rot = 90,
    col = c('white', 'tomato'),
    heatmap_legend_param = list(
    title = "Frequency",
    direction = "horizontal",
    title_gp = gpar(fontsize = 14, fontface = "bold"),
    labels_gp = gpar(fontsize = 14),
    legend_height = unit(5, "cm"),
    legend_width = unit(1.2, "cm")
    ))
```

## PRPS for TCGA and GSE81538
### UMAP plots of CCs
```{r}
keep.samples <- ms.se.obj$studies %in% c('GSE96058', 'GSE81538')
new.se.obj <- ms.se.obj[ , keep.samples]
set.seed(1133)
new.se.obj <- RUVprps:::computeCCA(
    se.obj        = new.se.obj, 
    assay.name    = 'FPKM', 
    apply.log     = FALSE,  
    normalization = NULL,
    nb.cca        = 30, 
    variable      = 'studies', 
    hvg           = hvg
    )
cca.corr <- new.se.obj@metadata$CCA$FPKM$studies$`Cca:30|HVG:13665|Scale:TRUE|Norm:`$GSE81538_GSE96058
cca.corr <- cca.corr[colnames(new.se.obj) , ]

set.seed(1133)
umap.all <- uwot::umap(
    cca.corr, 
    n_neighbors = 30, 
    min_dist    = 0.3, 
    metric      = "cosine"
    )
# plotting
data.frame(
  umap1            = umap.all[, 1],
  umap2            = umap.all[, 2],
  group            = new.se.obj$pam50) %>%
  ggplot(aes(x     = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = selectColors(
    nb.color = 1:5,
    group = 'pan.selection.a'
  )) +
  theme(
    panel.background = element_blank(),
    axis.line        = element_line(colour = 'black', linewidth = 1),
    axis.title.x     = element_text(size = 16),
    axis.title.y     = element_text(size = 16),
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 12),
    legend.title     = element_blank(),
    legend.text      = element_text(size = 16)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)), ncol = 2)   # enlarge legend points

data.frame(
  umap1 = umap.all[, 1],
  umap2 = umap.all[, 2],
  group = new.se.obj$studies) %>%
  ggplot(aes(x = umap1, y = umap2, color = group)) +
  geom_point(alpha = .4, size = 2, shape = 19, stroke = NA) +         # tiny points on plot
  xlab('UMAP1') +
  ylab('UMAP2') +
  scale_color_manual(values = c("GSE81538" = 'darkgreen', "GSE96058" = 'tomato', "TCGA" = 'darkblue')) +
  theme(
    panel.background = element_blank(),
    axis.line = element_line(colour = 'black', linewidth = 1),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 16)
  ) +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)), ncol = 2)   # enlarge legend points
```

### PAM50 distribution
```{r}
tbl <- table(new.se.obj$studies, new.se.obj$pam50)
ht <- ComplexHeatmap::Heatmap(
    matrix               = tbl , 
    name                 = 'Frequency',
    cluster_columns      = F,
    cluster_rows         = F,
    show_row_dend        = F, 
    show_column_dend     = F,
    column_names_rot     = 45,
    col                  = c('white', 'tomato'),
    heatmap_legend_param = list(
    title                = "Frequency",
    direction            = "horizontal",
    title_gp             = gpar(fontsize = 14, fontface = "bold"),
    labels_gp            = gpar(fontsize = 14),
    legend_height        = unit(5, "cm"),
    legend_width         = unit(1.2, "cm")))
ComplexHeatmap::draw(ht, heatmap_legend_side = "bottom")
```

### Create PRPS data
```{r}
new.se.obj <- createPrPsUnSupervisedByCca(
    se.obj             = new.se.obj, 
    assay.name         = 'FPKM',
    main.uv.variable   = 'studies',
    hvg                = hvg, 
    coordinates.to.use = "cca",
    nb.cca             = 30, 
    nb.pcs             = 3,
    apply.log          = F, 
    min.nb.for.mnn     = 5,
    apply.log.for.prps = F, 
    normalization      = NULL,
    max.prps.sets      = 20
    )
mnn          <- new.se.obj@metadata$PRPS$un.supervised$studies$`studies|CcaPca|FPKM`$mnn.samples
dim(mnn) # 50 4
sample.annot <- as.data.frame(colData(ms.se.obj))
a            <- sample.annot[mnn$left.samples, ]$pam50
b            <- sample.annot[mnn$right.samples, ]$pam50
tbl          <- table(a, b)

ComplexHeatmap::Heatmap(
    matrix               = tbl , 
    name                 = 'Frequency',
    cluster_columns      = F,
    cluster_rows         = F,
    show_row_dend        = F, 
    show_column_dend     = F,
    column_names_rot     = 90,
    col                  = c('white', 'tomato'),
    heatmap_legend_param = list(
    title                = "Frequency",
    direction            = "horizontal",
    title_gp             = gpar(fontsize = 14, fontface = "bold"),
    labels_gp            = gpar(fontsize = 14),
    legend_height        = unit(5, "cm"),
    legend_width         = unit(1.2, "cm")
    ))
```


